---
title: "sis-HMRF: Spatial Immune State Inference in BRCA Visium"
author: "Tao Zhu"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{sis-HMRF: BRCA Visium Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 6,
  fig.height = 5,
  message    = FALSE,
  warning    = FALSE
)
```

## Overview

The `sisHMRF` package implements a continuous-latent-variable hidden Markov
random field for inferring immune state landscapes from spatial transcriptomics
data. Unlike discrete-domain HMRF methods (BayesSpace, SC-MEB) that partition
tissue into sharp clusters, sis-HMRF models immune archetypes as continuous
latent variables under a Gaussian Markov random field (GMRF) prior.

This vignette demonstrates the core workflow using a 10x Visium breast cancer
(BRCA) dataset.

## Installation

```r
# From source
install.packages("sisHMRF_0.1.0.tar.gz", repos = NULL, type = "source")

# Dependencies
install.packages(c("Matrix", "FNN", "ggplot2", "scales", "reshape2", "patchwork"))
```

## Core workflow

### Step 1: Load spatial transcriptomics data

The package includes a pre-computed immune feature matrix from a BRCA Visium
sample (500 spots subset). For users with their own Visium data, see the
**Preparing your own data** section below.

```{r load-example}
library(sisHMRF)

# Load bundled BRCA example dataset
brca_path <- system.file("extdata", "brca_example.rds", package = "sisHMRF")
brca <- readRDS(brca_path)

cat(sprintf("Spots: %d  |  Features: %d\n",
            brca$n_spots, ncol(brca$F_mat)))
cat(sprintf("Feature categories:\n"))
print(table(brca$feature_categories))
```

### Step 2: Build immune feature matrix

For the bundled example, features are pre-computed. For your own expression
matrix, use `sis_features()`:

```{r features, eval=FALSE}
# expr_mat: N×G normalized expression matrix
features <- sis_features(expr_mat, verbose = TRUE)
print(features)

# Inspect feature matrix
dim(features$F_mat)
head(features$feature_names)
table(features$feature_categories)
```

### Step 3: Fit sis-HMRF model

```{r fit}
fit <- sis_fit(
  F_mat   = brca$F_mat,
  coords  = brca$coords,
  P       = 3,       # 3 latent immune dimensions
  k       = 6,       # 6 nearest neighbors
  verbose = FALSE
)

# Model summary
print(fit)
```

```{r fit-details}
summary(fit)
```

### Step 4: Visualize results

**Latent immune dimensions** — each panel shows one dimension's spatial
distribution across the tissue.

```{r vis-panel, fig.height=4, fig.width=9}
plot(fit, brca$coords, type = "panel")
```

**Immune scores** — IAS (activation), ISS (suppression), IBI (balance index).

```{r vis-scores, fig.height=3.5, fig.width=9}
plot(fit, brca$coords, type = "scores")
```

**Immune gradient** — continuous transition zones between immune activation and
suppression.

```{r vis-gradient, fig.height=5, fig.width=6}
plot(fit, brca$coords, type = "gradient")
```

**Post-hoc immune archetypes** — k-means clustering of latent dimensions reveals
discrete immune states (IS1--IS4).

```{r vis-archetypes, fig.height=5, fig.width=7}
plot(fit, brca$coords, type = "archetypes", K = 4)
```

**Feature loadings** — which immune features drive each latent dimension.

```{r vis-loadings, fig.height=6, fig.width=6}
plot(fit, brca$coords, type = "loadings")
```

**ELBO convergence** — verify model optimization.

```{r vis-elbo, fig.height=4, fig.width=5}
plot(fit, brca$coords, type = "elbo")
```

### Step 5: Extract immune scores

```{r scores}
scores <- sis_scores(fit)
cat("Top 6 spots by immune balance index (IBI):\n")
high_ibi <- order(scores[, "IBI"], decreasing = TRUE)
print(head(scores[high_ibi, ]))

cat(sprintf("\nIAS range: [%.3f, %.3f]\n", min(scores[,1]), max(scores[,1])))
cat(sprintf("ISS range: [%.3f, %.3f]\n", min(scores[,2]), max(scores[,2])))
cat(sprintf("IBI range: [%.3f, %.3f]\n", min(scores[,3]), max(scores[,3])))
```

### Step 6: Model selection (optional)

```{r model-selection, eval=FALSE}
bic_results <- sis_select_P(brca$F_mat, brca$coords, P_max = 8, k = 6)
print(bic_results)

# Plot BIC
plot(bic_results$P, bic_results$BIC, type = "b",
     xlab = "P (latent dimensions)", ylab = "BIC",
     main = "Model Selection via BIC")
```

## Preparing your own data

If you have raw 10x Visium data (Space Ranger output):

```{r own-data, eval=FALSE}
# Using Seurat to load Visium data
library(Seurat)
seurat_obj <- Load10X_Spatial("path/to/visium/sample")
count_matrix <- GetAssayData(seurat_obj, slot = "counts")
coords <- GetTissueCoordinates(seurat_obj)

# Normalize
expr_mat <- log1p(sweep(count_matrix, 1, rowSums(count_matrix) / 1e4, "/"))

# Build immune features
features <- sis_features(expr_mat)

# Fit model
fit <- sis_fit(features$F_mat, coords, P = 3, k = 6)

# Visualize
plot(fit, coords, type = "panel")
plot(fit, coords, type = "scores")
```

## Key parameters

| Parameter | Description | Default |
|-----------|-------------|---------|
| `P` | Number of latent immune dimensions | 3 |
| `k` | Spatial neighbors for graph | 6 |
| `max_iter` | Maximum EM iterations | 500 |
| `tol` | Convergence tolerance (ELBO relative change) | 1e-6 |
| `init_method` | Initialization: "pca" or "random" | "pca" |

## Output scores

| Score | Name | Interpretation |
|-------|------|----------------|
| IAS | Immune Activation Score | Higher = more immune activation |
| ISS | Immune Suppression Score | Higher = more immune suppression |
| IBI | Immune Balance Index | 0 = suppression-dominant, 1 = activation-dominant |

## Session info

```{r session-info}
sessionInfo()
```
